Manuscript Figures

### 2021-08-25 src/manuscript_figures.Rmd
Merge fig3 with fig4, survival-related LF analysis is now part of clinical factor analysis Drop Fig4B/C (example LFs associated to clinical factors), move 4A (clinical factor-associated LFs) to merge with Fig3A (cindex improvement using LFs), drop Fig3B (hazard ratio plot), move Fig3C (survival adjusted curves) to supplements

```{r survival_factors, fig.width=12, fig.height=8}
# survival-predictive factors
#I look for survival-predictive maui LFs, show how much c-index improves on top of 
#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
  require(data.table)
  require(survminer)
  require(survival)
  source(settings$utility_script)
  message(experiment)
  M <- LFs$maui_factors[[experiment]]
  # get clinical factors for the patients in matrix
  clin <- process_clinical_covariates(rownames(M))
  clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
  # first filter variables by importance 
  fit <- fit_RSF(surv, M, 500, time, event)
  vars <- names(head(sort(fit$importance, decreasing = T), 5))

  # split variable into high/low to compute survival adjusted curves
  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
  dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
  cox2 <- coxph(Surv(time,status) ~ ., 
                data=dat, x = TRUE)
  p <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')

  return(list('p' = p, 
              'Cindex_base' = rsf_stats$C_wo, 
              'Cindex_with_LFs' = rsf_stats$C_with))
  return(NULL)
})
parallel::stopCluster(cl)

# combine adjusted survival curve plots into a single plot
plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)

p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

Fig <- cowplot::plot_grid(cowplot::plot_grid(p3, p2, nrow = 1, labels = c('A', 'B')),
                        p1, labels = c('', 'C'), nrow = 2)

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 12, height = 14, units = 'in')
Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction. A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs. B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender). C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
```

```{r clinical_factors, fig.width=12, fig.height=9}
# clinical factor associated MAUI LFs
# Find maui LFs associated to clinical factors and plot them

@@

# pick example cases and show more detailed LF activation 
# top examples by clinical factor type
# top 5 variables per clinical factor 
top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
plots <- sapply(simplify = FALSE, 1:4, function(i) {
  pr <- top$project[i]
  l <- top$label[i]
  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
  factors <- process_clinical_covariates(rownames(M))[[l]]
  M <- M[!is.na(factors),]
  factors <- factors[!is.na(factors)]
  plot_label_specific_vars(M, factors, string_width = 24)+ 
    labs(title = pr, subtitle = l) + 
    theme(legend.position = 'none')
})

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))

Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.clinical_factors.LFs.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 12, height = 9)
Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type. B) Example case: top LFs associated with each gender groups of TCGA-THCA cohort. C) Example case: top LFs associated with each histological types of TCGA-UCEC."
```

>>>>

```{r clinical_survival_factors, fig.width=12, fig.height=8}
# here we compute LFs associated to clinical factors (e.g. tumor stage, gender etc) and 
# survival outcomes (PFI)

## 1. clinical factor associated MAUI LFs
# Find maui LFs associated to clinical factors and plot them

@@

# 2. survival-predictive factors
#I look for survival-predictive maui LFs, show how much c-index improves on top of 
#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) { 
  require(data.table)
  require(survminer)
  require(survival)
  source(settings$utility_script)
  message(experiment)
  M <- LFs$maui_factors[[experiment]]
  # get clinical factors for the patients in matrix
  clin <- process_clinical_covariates(rownames(M))
  clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
  # first filter variables by importance 
  fit <- fit_RSF(surv, M, 500, time, event)
  vars <- names(head(sort(fit$importance, decreasing = T), 5))

  # split variable into high/low to compute survival adjusted curves
  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
  dat[,v] <- ifelse(dat[,v] > cp, paste(v, '>', round(cp, 1)), paste(v, '<=', round(cp, 1)))
  cox2 <- coxph(Surv(time,status) ~ ., 
                data=dat, x = TRUE)
  p <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')

  return(list('p' = p, 
              'Cindex_base' = rsf_stats$C_wo, 
              'Cindex_with_LFs' = rsf_stats$C_with))
})
parallel::stopCluster(cl)

p2 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, nrow = 2, labels = c('A', 'B')))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 8.27, height = 11.69, units = 'in')
Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction. A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs. B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender). C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."

# combine adjusted survival curve plots into a single plot 
plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p3 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)

# print p3 to supplements 
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- p3
fig.name <- paste0('SuppFigure', supFig_count, '.adjusted_survival_curves.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 8.27, height = 11.69)
SuppLegends[[supFig_count]] <- "Survival curves adjusted for tumor stage + age + gender based on a cutpoint of 
top survival-associated LF per cancer cohort."
